Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-11-23 concurrency/CPP 2024/paper.tex
a bit hack for traverse spec

We prove correctness of the template by showing that \lstinline{traverse}, \lstinline{insert}, etc. meet logically atomic specifications describing their effects on the data structure. These specifications are defined in terms of assertions $\nodeboxrep$, representing a client thread's handle to the data structure, and $\treerep$, representing the data structure's abstract state. Holding $\nodeboxrep(n)$ for a node $n$ should allow us to acquire the lock on $n$, which in turn gives us access to the contents of the node. Formally, we need to know that the following triples hold:

@@

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
{ \color{blue}
	\fforall \  m. \ \scaleleftright[1.50ex]{\langle}
	{\begin{array}{l}
			\texttt{pn} \mapsto (p, n) \ \ast \ 
			\nodeboxrep(n)  \ast \ \mathsf{R}(n) 
		\end{array}  
		\ \Bigg \vert  \treerep\ (m)
	}  
	{\rangle}
}
\vspace{-6pt}\\
\vspace{-6pt}\texttt{traverse(pn, k)} \\
{ \color{blue}
\scaleleftright[1.50ex]{\langle}
{\begin{array}{l}
		\mathit{res}. \ \exists \  n', v, \mathsf{range}. \\ 
		\ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		\ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ 
		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right)
\end{array}  
\Bigg \vert \treerep\ (m)
}  
{\rangle}
}
\end{mathpar}

@@

The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure? Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations. However, in most sequential settings ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself. The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to data structures without modifying them. Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t}:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {
		node *t; 
		lock_t *lock; 
		int min, max;
	} node_t;
\end{lstlisting}
These fields store lower and upper bounds on the keys reachable from the current node. %In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it. %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
Our specification of \lstinline{inRange} is:
\begin{mathpar}
	% inRange for give-up
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2
		\end{array}
		\right\}
	} \vspace{-0.85em} 
	\\ \texttt{inRange(pn, k)} 
	\vspace{-0.85em}  \\
	{\color{blue}
		\left\{\mathit{res.} \ 
		\begin{array}{l}
			\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2 \ \ast  \ 
			\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2))
		\end{array}
		\right\}
	}
\end{mathpar}
It simply computes whether the input key \lstinline{k} is in the range $(n_1, n_2)$ associated with the node \texttt{n}, and returns 0 or 1 accordingly.

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{\color{blue}
		\fforall \  m. \left\langle
		\begin{array}{c}
			\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ (m)
		\end{array}
		\right\rangle
	}
	\vspace{-6pt}\\ \vspace{-6pt}
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \\ \texttt{k} \in \mathsf{range}\ \ast\ 
			\\ 
			(\mathsf{if} \ \mathit{res} \\ 
			\ \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ 
			\ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
		\end{array}
		\Bigg| \ \treerep\ (m) \
		\right\rangle
	}
\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}. Our definition of $\treerep$ for the give-up template closely follows that of Krishna et al.

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
\end{align*}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.

@@

\begin{figure*}[!ht]
	\centering
	\begin{mathpar}
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)
			\end{array}
			\right\}
		} \vspace{-0.85em} 
		\\ 
		\texttt{findNext(pn, k)} 
		\vspace{-0.85em} 
		\\
		{\color{blue}
			\left\{\mathit{res}. \ \exists \  n'.
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)  \ \ast \\\  
				(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k'))
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)}
	\label{fig:findNext_lock}
\end{figure*}

\begin{figure*}[h]
	\begin{subfigure}{\textwidth}
		\centering
		\begin{mathpar}
			{\color{blue}
				\left\{ 
				\begin{array}{c}
					\ \texttt{pn->n} \mapsto \texttt{NULL}  
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\ 
			\texttt{insertOp(pn, k, v)} 
			\vspace{-0.85em} 
			\\
			{\color{blue}
				\left\{\exists \  l\ r\ \mathit{lk}_1\ \mathit{lk}_2.
				\begin{array}{l}
					\ \texttt{pn->n->t} \mapsto (\texttt{k}, \texttt{v}, l, r) \ \ast \ \\ 
					\ l \mapsto (\texttt{NULL}, \mathit{lk}_1)\  \ast \ r \mapsto (\texttt{NULL}, \mathit{lk}_2)
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the lock-coupling template}
		\label{fig:insertOp_lock}
	\end{subfigure}
	%insertOp for give-up of give-up
	\begin{subfigure}{\textwidth}
		\begin{mathpar}
			{\color{blue}
				\left\{ 
				\begin{array}{c}
					\texttt{pn->n}\mapsto \texttt{NULL} \  \ast \ \texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\
			\texttt{insertOp(pn, k, v)}
			\vspace{-0.85em} 
			\\
			{\color{blue}
				\left\{\exists \ l\ r\ \mathit{lk}_1\ \mathit{lk}_2. 
				\begin{array}{c}
					\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r) \  \ast \ \texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2  \ \ast \ 
					\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (n_1, \texttt{k})) \ \ast \ r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, n_2))
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the give-up template}
		\label{fig:insertOp_giveup}
	\end{subfigure}
	\caption{Specification of \texttt{insertOp} for the lock-coupling and give-up templates}
	\label{fig:insertOp}
\end{figure*}

@@

\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ (m)
	\end{array}
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
	\end{array}
	\ \Bigg| \ \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the give-up \texttt{traverse} function}
	\label{proof_giveup_traverse}
\end{figure*}

>>>>

We prove correctness of the template by showing that \lstinline{traverse}, \\ \lstinline{insert}, etc. meet logically atomic specifications describing their effects on the data structure. These specifications are defined in terms of assertions $\nodeboxrep$, representing a client thread's handle to the data structure, and $\treerep$, representing the data structure's abstract state. Holding $\nodeboxrep(n)$ for a node $n$ should allow us to acquire the lock on $n$, which in turn gives us access to the contents of the node. Formally, we need to know that the following triples hold:

@@

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
{ \color{blue}
	\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
	{\begin{array}{l}
			\texttt{pn} \mapsto (p, n) \ \ast \ 
			\nodeboxrep(n)  \ast \ \mathsf{R}(n) 
		\end{array}  
		\Big \vert  \treerep\ (m)
	}  
	{\rangle}
}
\vspace{-6pt}\\
\vspace{-6pt}\texttt{traverse(pn, k)} \\
{ \color{blue}
 \scaleleftright[1.4ex]{\langle}
{\begin{array}{l}
		\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
		\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ 
		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
\end{array}  
\Bigg \vert \treerep\ (m)
}  
{\rangle}
}
\end{mathpar}

@@

The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure? Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations. However, in most sequential settings ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself. The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to data structures without modifying them. Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t}:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {
		node *t; 
		lock_t *lock; 
		int min, max;
	} node_t;
\end{lstlisting}
These fields store lower and upper bounds on the keys reachable from the current node. %In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it. %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
Our specification of \lstinline{inRange} is:
\begin{mathpar}
	% inRange for give-up
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2
		\end{array}
		\right\}
	} \vspace{-0.85em} 
	\\ \texttt{inRange(pn, k)} 
	\vspace{-0.85em}  \\
	{\color{blue}
		\left\{\mathit{res.} 
		\begin{array}{l}
			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2 \ \ast  \ 
			\\\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2)
		\end{array}
		\right\}
	}
\end{mathpar}
It simply computes whether the input key \lstinline{k} is in the range $(n_1, n_2)$ associated with the node \texttt{n}, and returns 0 or 1 accordingly.

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{ \color{blue}
		\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\texttt{pn} \mapsto (p, n) \ \ast \ 
				\nodeboxrep(n)   
			\end{array}  
			\Big \vert  \treerep\ (m)
		}  
		{\rangle}
	}
	\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{ \color{blue}
		\scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range} \ \ast \ 
				\\ 
				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ 
				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}  
			\Bigg \vert \treerep\ (m)
		}  
		{\rangle}
	}
\end{mathpar}

As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}. Our definition of $\treerep$ for the give-up template closely follows that of Krishna et al.

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast  \infp (r) \ast \infp (n)
\end{align*}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.

@@

\begin{figure*}[!ht]
	\centering
	\begin{mathpar}
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)
			\end{array}
			\right\}
		} \vspace{-0.85em} 
		\\ 
		\texttt{findNext(pn, k)} 
		\vspace{-0.85em} 
		\\
		{\color{blue}
			\left\{\mathit{res}. \ \exists \  n'.
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)  \ \ast \\\  
				\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k')
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)}
	\label{fig:findNext_lock}
\end{figure*}

\begin{figure*}[h]
	\begin{subfigure}{\textwidth}
		\centering
		\begin{mathpar}
			{\color{blue}
				\left\{ 
				\begin{array}{c}
					\ \texttt{pn->n} \mapsto \texttt{NULL}  
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\ 
			\texttt{insertOp(pn, k, v)} 
			\vspace{-0.85em} 
			\\
			{\color{blue}
				\left\{\exists \  l\ r\ \mathit{lk}_1\ \mathit{lk}_2.
				\begin{array}{l}
					\ \texttt{pn->n->t} \mapsto (\texttt{k}, \texttt{v}, l, r) \ \ast \ \\ 
					\ l \mapsto (\texttt{NULL}, \mathit{lk}_1)  \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2)
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the lock-coupling template}
		\label{fig:insertOp_lock}
	\end{subfigure}
	%insertOp for give-up of give-up
	\begin{subfigure}{\textwidth}
		\begin{mathpar}
			{\color{blue}
				\left\{ 
				\begin{array}{c}
					\texttt{pn->n}\mapsto \texttt{NULL}  \ast  \texttt{pn->n->min}\mapsto n_1  \ast  \texttt{pn->n->max}\mapsto n_2
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\
			\texttt{insertOp(pn, k, v)}
			\vspace{-0.85em} 
			\\
			{\color{blue}
				\left\{\exists \ l\ r\ \mathit{lk}_1\ \mathit{lk}_2. 
				\begin{array}{c}
					\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r)  \ast  \texttt{pn->n->min}\mapsto n_1 \ast  \texttt{pn->n->max}\mapsto n_2  \ \ast \ 
					\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (n_1, \texttt{k})) \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, n_2))
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the give-up template}
		\label{fig:insertOp_giveup}
	\end{subfigure}
	\caption{Specification of \texttt{insertOp} for the lock-coupling and give-up templates}
	\label{fig:insertOp}
\end{figure*}

@@

\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (p, n)  \ast  \nodeboxrep(n)  \big| \ \treerep\ (m)
	\end{array}
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
	\end{array}
	\ \Bigg| \ \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the give-up \texttt{traverse} function}
	\label{proof_giveup_traverse}
\end{figure*}
